---
title: "AI Hardware Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Install required packages if missing:
# install.packages(c("tidyverse", "lubridate", "janitor"))

library(tidyverse)
library(lubridate)
library(stringr)
library(janitor)
```

# Data Processing

To accurately synthesize this data, a many-to-one relational join is required. This function standardizes names across all datasets to ensure clean joins.

```{r}
clean_gpu_string <- function(name_col) {
  name_col %>%
    str_to_lower() %>%
    str_remove_all("msi|asus|gigabyte|zotac|evga|sapphire|powercolor|pny") %>%
    str_remove_all("nvidia|amd|geforce|radeon|founders edition") %>%
    str_squish()
}
```

We load the five disparate datasets required for the high-dimensional dataset.
```{r}
# TODO: filter out columns/rows keeping only those relevant to the analysis, e.g., retail price, GPU name, benchmark scores, etc.
# Assuming files are in the working directory
video_cards <- read_csv("data/video-card.csv") %>% clean_names()
gpu_benchmarks <- read_csv("data/GPU_benchmarks_v7.csv") %>% clean_names()
ml_hardware <- read_csv("data/ml_hardware.csv") %>% clean_names()
shs <- read_csv("data/shs.csv") %>% clean_names()
ppi_data <- read_csv("data/PCU334413334413.csv") %>% clean_names()
```

We map the chipset column in the video-card.csv data directly to the gpuName column in GPU_benchmarks_v7.csv.

```{r}
# Process Retail Video Cards (video-card.csv)
video_cards_clean <- video_cards %>%
  mutate(join_name = clean_gpu_string(chipset)) %>%
  group_by(join_name) %>%
  summarise(
    avg_retail_price = mean(price, na.rm = TRUE),
    vram_capacity = max(memory, na.rm = TRUE),
    core_clock = max(core_clock, na.rm = TRUE),
    boost_clock = max(boost_clock, na.rm = TRUE)
  )
# Process Traditional Benchmarks (GPU_benchmarks_v7.csv)
benchmarks_clean <- gpu_benchmarks %>%
  mutate(join_name = clean_gpu_string(gpu_name)) %>%
  select(join_name, category, g3dmark, g2dmark) %>%
  # Handle potential duplicates if generalized names overlap
  distinct(join_name, .keep_all = TRUE)
# Process AI Architectural Specs (ml_hardware.csv)
ml_hardware_clean <- ml_hardware %>%
  mutate(join_name = clean_gpu_string(hardware_name)) %>%
  select(
    join_name,
    fp16_flops = fp16_half_precision_performance_flop_s,
    fp32_flops = fp32_single_precision_performance_flop_s,
    fp64_flops = fp64_double_precision_performance_flop_s,
    tensor_cores,
    memory_bandwidth = memory_bandwidth_byte_s
  ) %>%
  distinct(join_name, .keep_all = TRUE)
# Merge Static Dictionary
hardware_dictionary <- video_cards_clean %>%
  inner_join(benchmarks_clean, by = "join_name") %>%
  left_join(ml_hardware_clean, by = "join_name") %>%
  drop_na(avg_retail_price) # Filter to actionable retail data
```

We process the Steam Hardware Survey for lagged consumer adoption and the FRED PPI for inflation control.

```{r}
# Process Steam Hardware Survey (shs.csv)
shs_clean <- shs %>%
  mutate(
    date = ymd(date),
    month_year = floor_date(date, "month"),
    join_name = clean_gpu_string(name)
  ) %>%
  select(month_year, join_name, percentage)
# Process Baseline Inflation Control (PCU334413334413.csv)
ppi_clean <- ppi_data %>%
  rename(date = observation_date, semi_ppi = pcu334413334413) %>%
  mutate(
    date = ymd(date),
    month_year = floor_date(date, "month")
  ) %>%
  select(month_year, semi_ppi)
# Merge Time-Series Data
time_series_data <- shs_clean %>%
  left_join(ppi_clean, by = "month_year")
```

This data structure broadcasts the underlying silicon capabilities to all partner variations, enabling the statistical models to capture exact retail price variance.

```{r}
final_analytical_dataset <- time_series_data %>%
  inner_join(hardware_dictionary, by = "join_name") %>%
  mutate(
    # Baseline Inflation Control Engineering
    inflation_adjusted_price = avg_retail_price / (semi_ppi / 100),
    # Target Casting for Logistic Regression
    is_workstation = as.factor(ifelse(category == "Workstation", 1, 0)),
    # Ensure categorical features are appropriately typed
    category = as.factor(category)
  ) %>%
  select(-any_of(c("price_y", "price_x")))
# View the final structured data
glimpse(final_analytical_dataset)
```

```{r export-csv}
# Save the final analytical dataset to your working directory
write_csv(final_analytical_dataset, "data/final_analytical_dataset.csv")
```