---
title: "AI Hardware Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Install required packages if missing:
# install.packages(c("tidyverse", "lubridate", "janitor", "caret"))

library(tidyverse)
library(lubridate)
library(stringr)
library(janitor)
library(caret)
```

# Data Processing

To accurately synthesize this data, a many-to-one relational join is required. This function standardizes names across all datasets to ensure clean joins.

```{r}
clean_gpu_string <- function(name_col) {
  name_col %>%
    str_to_lower() %>%
    # Remove brand fluff
    str_remove_all("\\b(msi|asus|gigabyte|zotac|evga|sapphire|powercolor|pny)\\b") %>%
    str_remove_all("\\b(nvidia|amd|ati|founders|collectors|anniversary|frontier|edition|50th)\\b") %>%
    # Destroy hyphens followed by numbers (fixes the " - 512" issue)
    str_remove_all("-\\s*[0-9]+") %>%
    # Target VRAM (3gb), Wattage (75w), minor variants (g5, ddr4), and specific editions (se, oem, lhr, x2, sff)
    str_remove_all("\\b([0-9]+gb|[0-9]+w|g[56]|ddr[0-9]|se|oem|lhr|x2|sdi|sff)\\b") %>%
    # Remove the word "series" from SHS
    str_remove_all("\\bseries\\b") %>%
    # Squish multiple spaces down to a single space
    str_squish() %>%
    # Handle AMD Radeon HD groupings (e.g., changes "radeon hd 7750" to "radeon hd 7700")
    str_replace_all("radeon hd ([0-9]{2})[0-9]{2}[a-z]*", "radeon hd \\100") %>%
    # Handle AMD R7/R9 Series groupings (e.g., changes "r9 290x" to "r9 200")
    str_replace_all("r9 2[0-9]{2}[a-z]*", "r9 200") %>%
    str_replace_all("r9 3[0-9]{2}[a-z]*", "r9 300") %>%
    str_replace_all("r7 2[0-9]{2}[a-z]*", "r7 200") %>%
    str_replace_all("r7 3[0-9]{2}[a-z]*", "r7 300")
}
```

We load the five disparate datasets required for the high-dimensional dataset.

```{r}
# Assuming files are in the working directory
video_cards <- read_csv("data/video-card.csv") %>% clean_names()
gpu_benchmarks <- read_csv("data/GPU_benchmarks_v7.csv") %>% clean_names()
ml_hardware <- read_csv("data/ml_hardware.csv") %>% clean_names()
shs <- read_csv("data/shs.csv") %>% clean_names()
ppi_data <- read_csv("data/PCU334413334413.csv") %>% clean_names()
```

#### Process Retail Video Cards (video-card.csv)
Keeping NA/0 prices, dropping irrelevant string columns

```{r}
video_cards_clean <- video_cards %>%
  select(-name, -color, -length) %>%
  mutate(chipset = clean_gpu_string(chipset)) %>%
  group_by(chipset) %>%
  summarise(
    avg_retail_price = mean(price, na.rm = TRUE),
    vram_capacity = max(memory, na.rm = TRUE),
    core_clock = max(core_clock, na.rm = TRUE),
    boost_clock = max(boost_clock, na.rm = TRUE)
  ) %>%
  # Catch -Inf (from max) and NaN (from mean) and safely convert back to NA
  mutate(
    across(c(vram_capacity, core_clock, boost_clock), ~na_if(., -Inf)),
    avg_retail_price = na_if(avg_retail_price, NaN)
  ) %>%
  distinct(chipset, .keep_all = TRUE)
```

#### Process Traditional Benchmarks (GPU_benchmarks_v7.csv)
Keeping price, dropping mobile segments

```{r}
benchmarks_clean <- gpu_benchmarks %>%
  mutate(
    # Create a temporary lowercase name for easy regex searching
    lower_name = str_to_lower(gpu_name),
    # Intelligently reclassify 'Unknown' (and correct mislabeled) GPUs
    category = case_when(
      # If it has laptop terminology, force it to Mobile
      str_detect(lower_name, "mobile|max-q|laptop|notebook") ~ "Mobile",
      # If it has enterprise/AI terminology, force it to Workstation
      str_detect(lower_name, "quadro|tesla|firepro|radeon pro|rtx a[0-9]") ~ "Workstation",
      # If it's still 'Unknown' after the above checks, safely assume it's a Desktop consumer card
      category == "Unknown" ~ "Desktop",
      # Otherwise, keep the original valid classification
      TRUE ~ category
    )
  ) %>%
  # Now apply the strict filter with confidence that the data is accurately tagged
  filter(category %in% c("Desktop", "Workstation")) %>%
  mutate(
    chipset = clean_gpu_string(gpu_name),
    chipset = str_replace(chipset, "radeon rx 570x", "radeon rx 570")
  ) %>%
  select(chipset, category, g3dmark, g2dmark, tdp, benchmark_price = price) %>%
  distinct(chipset, .keep_all = TRUE)
```

#### Process Steam Hardware Survey (shs.csv)
Keeping 0 percentages, filtering out CPUs/OS metrics

```{r}
shs_clean <- shs %>%
  filter(category == "Video Card Description") %>%
  # Drop laptop chips (mobility/mobile) and integrated CPU graphics (intel)
  filter(!str_detect(str_to_lower(name), "mobility|mobile|intel")) %>%
  mutate(
    date = ymd(date),
    month_year = floor_date(date, "month"),
    chipset = clean_gpu_string(name)
  ) %>%
  # Apply strict timeline filter
  filter(month_year >= ymd("2009-01-01") & month_year <= ymd("2025-12-31")) %>%
  select(month_year, chipset, percentage)
```

#### Process AI Architectural Specs (ml_hardware.csv)
Filtering for GPUs only. Keeping all numeric specs, dropping irrelevant string columns

```{r}
ml_hardware_clean <- ml_hardware %>%
  filter(type == "GPU") %>%
  mutate(chipset = clean_gpu_string(hardware_name)) %>%
  select(
    chipset,
    ml_release_price = matches("release_price"),
    ml_memory = matches("memory_bytes"),
    ml_tdp = matches("tdp"),
    fp32_flops = matches("fp32_single"),
    ml_base_clock = matches("base_clock"),
    ml_boost_clock = matches("boost_clock"),
    tensor_cores,
  ) %>%
  mutate(
    across(
      c(ml_memory, fp32_flops),
      ~ .x / 1e9
    )
  ) %>%
  distinct(chipset, .keep_all = TRUE)
```

Add missing cards' informations manually based on producers' websites and Wikipedia

```{r}
missing_gpus <- tibble(
  chipset = c(
    "geforce rtx 5050", "geforce rtx 5060", "geforce rtx 5060 ti", "geforce rtx 5070",
    "geforce rtx 5070 ti", "geforce rtx 5080", "geforce rtx 5090",
    "rtx 4000 ada generation", "rtx 5000 ada generation",
    "radeon rx 9060 xt", "radeon rx 9070", "radeon rx 9070 xt",
    "arc b570", "arc b580", "arc a310", "arc a380", "arc a580", "arc a750", "arc a770",
    "geforce rtx 4060", "geforce rtx 4060 ti", "geforce rtx 4070", "geforce rtx 4070 super",
    "geforce rtx 4070 ti", "geforce rtx 4070 ti super", "geforce rtx 4080 super",
    "radeon rx 7600", "radeon rx 7600 xt", "radeon rx 7700 xt", "radeon rx 7800 xt",
    "radeon rx 7900 gre", "radeon rx 7900 xt",
    "geforce rtx 3050", "geforce rtx 3060", "geforce rtx 3060 ti", "geforce rtx 3070",
    "geforce rtx 3070 ti", "geforce rtx 3080", "geforce rtx 3080 ti", "geforce rtx 3090", "geforce rtx 3090 ti",
    "geforce rtx 2060", "geforce rtx 2060 super", "geforce rtx 2070", "geforce rtx 2070 super",
    "geforce rtx 2080", "geforce rtx 2080 super", "geforce rtx 2080 ti",
    "radeon rx 6650 xt", "radeon rx 6750 xt", "radeon rx 6950 xt",
    "rtx 2000 ada generation", "rtx 2000e ada generation", "rtx 4500 ada generation",
    "radeon pro w7500", "radeon pro w7600", "radeon pro w7700", "radeon pro w7800", "radeon pro w7900",
    "radeon pro w6600", "radeon pro w6800",
    "radeon rx 6500 xt", "radeon rx 6600", "radeon rx 6600 xt", "radeon rx 6700 xt",
    "radeon rx 6800", "radeon rx 6800 xt", "radeon rx 6900 xt"
  ),

  ml_release_price = c(
    249, 299, 429, 549, 749, 999, 1999,
    1250, 4000,
    399, 549, 599,
    219, 249, 110, 139, 179, 249, 329,
    299, 399, 599, 599, 799, 799, 999,
    269, 329, 449, 499, 549, 899,
    249, 329, 399, 499, 599, 699, 1199, 1499, 1999,
    349, 399, 499, 499, 699, 699, 999,
    399, 549, 1099,
    625, 625, 2250,
    429, 599, 999, 2499, 3999,
    649, 2249,
    199, 329, 379, 479, 579, 649, 999
  ),

  ml_memory = c(
    8, 8, 16, 12, 16, 16, 32,
    20, 32,
    16, 16, 16,
    10, 12, 4, 6, 8, 8, 16,
    8, 8, 12, 12, 12, 16, 16,
    8, 16, 12, 16, 16, 20,
    8, 12, 8, 8, 8, 10, 12, 24, 24,
    6, 8, 8, 8, 8, 8, 11,
    8, 12, 16,
    16, 16, 24,
    8, 8, 16, 32, 48,
    8, 32,
    4, 8, 8, 12, 16, 16, 16
  ),

  ml_tdp = c(
    130, 145, 180, 250, 300, 360, 575,
    130, 250,
    160, 220, 304,
    150, 190, 75, 75, 185, 225, 225,
    115, 160, 200, 220, 285, 285, 320,
    165, 190, 245, 263, 260, 300,
    130, 170, 200, 220, 290, 320, 350, 350, 450,
    160, 175, 175, 215, 215, 250, 250,
    180, 250, 335,
    70, 50, 210,
    70, 130, 190, 260, 295,
    130, 250,
    107, 132, 160, 230, 250, 300, 300
  ),

  fp32_flops = c(
    13.17, 19.18, 23.70, 30.84, 43.94, 56.28, 104.80,
    26.70, 65.30,
    25.64, 36.13, 48.66,
    12.70, 13.67, 3.07, 4.09, 10.42, 14.64, 17.20,
    15.11, 22.06, 29.15, 35.48, 40.09, 44.10, 52.22,
    21.75, 22.57, 35.17, 37.32, 43.01, 51.48,
    9.10, 12.74, 16.20, 20.31, 21.75, 29.77, 34.10, 35.58, 40.00,
    6.45, 7.18, 7.46, 9.06, 10.07, 11.15, 13.45,
    10.79, 13.31, 23.65,
    12.0, 8.9, 39.6,
    12.2, 16.6, 28.3, 45.3, 61.3,
    10.4, 26.7,
    5.77, 8.93, 10.60, 13.21, 16.17, 20.74, 23.04
  ),

  tensor_cores = c(
    80, 120, 144, 192, 280, 336, 680,
    192, 400,
    64, 112, 128,
    144, 160, 96, 128, 384, 448, 512,
    96, 136, 184, 224, 240, 264, 320,
    64, 64, 108, 120, 160, 168,
    80, 112, 152, 184, 192, 272, 320, 328, 336,
    240, 272, 288, 320, 368, 384, 544,
    32, 40, 80,
    88, 88, 240,
    28, 32, 48, 70, 96,
    28, 60,
    16, 28, 32, 40, 60, 72, 80
  )
)
# Bind the missing cards BEFORE the main dataset so our complete specs overwrite any NAs
ml_hardware_clean <- missing_gpus %>%
  bind_rows(ml_hardware_clean) %>%
  distinct(chipset, .keep_all = TRUE)
```

#### Process Baseline Inflation Control (FRED's PCU334413334413.csv)
Filtering out historical data that precedes the SHS timeline

```{r}
min_shs_date <- min(shs_clean$month_year, na.rm = TRUE)
ppi_clean <- ppi_data %>%
  rename(date = observation_date, semi_ppi = pcu334413334413) %>%
  mutate(
    date = ymd(date),
    month_year = floor_date(date, "month")
  ) %>%
  # Apply strict timeline filter
  filter(month_year >= ymd("2009-01-01") & month_year <= ymd("2025-12-31")) %>%
  select(month_year, semi_ppi)
```

Write the Clean Datasets to CSV

```{r}
write_csv(video_cards_clean, "data/video_cards_clean.csv")
write_csv(benchmarks_clean, "data/benchmarks_clean.csv")
write_csv(shs_clean, "data/shs_clean.csv")
write_csv(ml_hardware_clean, "data/ml_hardware_clean.csv")
write_csv(ppi_clean, "data/ppi_clean.csv")
```

#### Prepare and export the GPUs Dictionary

```{r}
# Full Join
hardware_dictionary <- video_cards_clean %>%
  full_join(benchmarks_clean, by = "chipset") %>%
  full_join(ml_hardware_clean, by = "chipset") %>%
# Category Imputation
  mutate(
    # Fixes NA categories for newer cards not in the benchmark dataset
    category = case_when(
      !is.na(category) ~ category, # Keep existing valid categories
      str_detect(chipset, "geforce rtx|gtx|rx|radeon hd|arc a|arc b|geforce") ~ "Desktop", # Consumer gaming tags
      str_detect(chipset, "pro|quadro|tesla|instinct") ~ "Workstation", # Enterprise tags
      TRUE ~ "Workstation" # Default remaining unknowns to Workstation
    ),
# Consolidate Overlapping Columns (Averaging)
    # Convert all 0 values to NA across overlapping columns
    across(
      c(avg_retail_price, benchmark_price, ml_release_price,
        core_clock, ml_base_clock,
        boost_clock, ml_boost_clock,
        tdp, ml_tdp,
        vram_capacity, ml_memory),
      ~na_if(., 0)
    ),
    # Calculate Unified Means (suppressWarnings hides NaN warnings if all are NA)
    unified_price = suppressWarnings(rowMeans(cbind(avg_retail_price, benchmark_price, ml_release_price), na.rm = TRUE)),
    unified_base_clock = suppressWarnings(rowMeans(cbind(core_clock, ml_base_clock), na.rm = TRUE)),
    unified_boost_clock = suppressWarnings(rowMeans(cbind(boost_clock, ml_boost_clock), na.rm = TRUE)),
    unified_tdp = suppressWarnings(rowMeans(cbind(tdp, ml_tdp), na.rm = TRUE)),
    unified_memory = suppressWarnings(rowMeans(cbind(vram_capacity, ml_memory), na.rm = TRUE))
  ) %>%
  # Convert any mathematical NaN results back to proper NA
  mutate(across(starts_with("unified_"), ~na_if(., NaN))) %>%
  # Drop the original redundant columns
  select(
    -avg_retail_price, -benchmark_price, -ml_release_price,
    -core_clock, -ml_base_clock,
    -boost_clock, -ml_boost_clock,
    -tdp, -ml_tdp,
    -vram_capacity, -ml_memory
  )
```

#### Regression Imputation for Missing Benchmarks



```{r}
# Train a simple model on GPUs that have both g3dmark and fp32_flops
g3d_model <- lm(g3dmark ~ fp32_flops, data = filter(hardware_dictionary, !is.na(g3dmark) & !is.na(fp32_flops)))
g2d_model <- lm(g2dmark ~ fp32_flops, data = filter(hardware_dictionary, !is.na(g2dmark) & !is.na(fp32_flops)))

hardware_dictionary <- hardware_dictionary %>%
  mutate(
    predicted_g3d = predict(g3d_model, newdata = .),
    predicted_g2d = predict(g2d_model, newdata = .),
    # If g3dmark or g2dmark are missing (new cards), fill it with the prediction based on FLOPs
    g3dmark = case_when(
      is.na(g3dmark) & !is.na(fp32_flops) ~ predicted_g3d,
      TRUE ~ g3dmark
    ),
    g2dmark = case_when(
      is.na(g2dmark) & !is.na(fp32_flops) ~ predicted_g2d,
      TRUE ~ g2dmark
    )
  ) %>%
  # Drop the temporary prediction columns
  select(-predicted_g3d, -predicted_g2d)

# Structural Zeros & Final Clean Up
hardware_dictionary <- hardware_dictionary %>%
  mutate(
    # Fix Tensor Cores for pre-2018 cards
    tensor_cores = case_when(
      !is.na(tensor_cores) ~ tensor_cores,
      str_detect(chipset, "gtx|gt |radeon hd|r7|r9|rx 4|rx 5|firepro|quadro k|quadro m|quadro p|titan x|titan z|titan xp|vega|radeon vii|tesla|p40|p100|m40|radeon pro duo|radeon pro vii|radeon pro ssg|radeon pro w5|radeon pro wx") ~ 0,
      g3dmark < 8000 & is.na(tensor_cores) ~ 0,
      TRUE ~ tensor_cores
    )
  ) %>%
  # THE CRITICAL FILTER: Drop rows where the final price is entirely missing
  drop_na(unified_price) %>%
  # Grouped median imputation to rescue slightly missing Memory and TDP values
  group_by(category) %>%
  mutate(
    unified_memory = ifelse(is.na(unified_memory), median(unified_memory, na.rm = TRUE), unified_memory),
    unified_tdp = ifelse(is.na(unified_tdp), median(unified_tdp, na.rm = TRUE), unified_tdp)
  ) %>%
  ungroup() %>%
  # Cast classification target
  mutate(
    category = as.factor(category)
  ) %>%
  # Drop unified clock speeds (they are highly missing and redundant now that g3dmark is populated)
  select(-unified_base_clock, -unified_boost_clock) %>%
  mutate(
    # Round and force structural/performance metrics to be integers
    across(
      c(g3dmark, g2dmark, tensor_cores, unified_tdp, unified_memory),
      ~ as.integer(round(.))
    ),
    # Ensure price remains strictly a floating point (double/numeric)
    unified_price = round(as.numeric(unified_price), 2)
  ) %>%
  mutate(
    tensor_cores = case_when(
      chipset == "quadro gv100" ~ 640,
      chipset == "rtx 6000 ada generation" ~ 568,
      chipset == "rtx a2000" ~ 104,
      chipset == "rtx a4500" ~ 224,
      chipset == "titan rtx" ~ 576,
      chipset == "instinct mi300x" ~ 1216,
      chipset == "radeon instinct mi250x" ~ 880,
      chipset == "hgx h20" ~ 456,
      chipset == "huawei ascend 910b" ~ 25,
      TRUE ~ tensor_cores
    )
  ) %>%
  # Drop fp32_flops (to prevent perfect collinearity in the ISLR models)
  select(-fp32_flops)

write_csv(hardware_dictionary, "data/hardware_dictionary_static.csv", na = "NA")
```

# Missing Data Analysis

#### Column-Level Missingness: Which features are missing the most?

```{r}
missing_cols <- hardware_dictionary %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Feature", values_to = "Missing_Count") %>%
  mutate(Missing_Percentage = (Missing_Count / nrow(hardware_dictionary)) * 100) %>%
  arrange(desc(Missing_Count))

print(missing_cols)
```

All the above results belong to workstation gpus and a notebook one. The notebook card will be dropped per analysis scope, while regarding the other four, since they also have the fp32 flops value missing, we can't predict the performance marks. Thus, considering the missing data is so low, we will drop these cards from the dataset.

#### Row-Level Missingness: Which specific GPUs are missing the most data?

```{r}
missing_rows <- hardware_dictionary %>%
  mutate(total_missing_specs = rowSums(is.na(.))) %>%
  filter(total_missing_specs > 0) %>% # Only keep rows that actually have NAs
  select(chipset, category, total_missing_specs) %>%
  arrange(desc(total_missing_specs))

print(missing_rows)
```

The above table shows which GPUs are missing specifications and will be dropped: we see the "radeon rx 6600m" card that only misses the tensor cores value, but as discussed above, it is a notebook card (misclassified as a Desktop card) that will be dropped.

# Final Data Preparation

#### Drop GPUs with Missing Specifications

```{r}
hardware_dictionary <- hardware_dictionary %>%
  drop_na()
hardware_dictionary_shs_only <- hardware_dictionary %>%
  semi_join(shs_clean, by = "chipset")

write_csv(hardware_dictionary, "data/hardware_dictionary_static.csv", na = "NA")
write_csv(hardware_dictionary_shs_only, "data/hardware_dictionary_shs_only.csv", na = "NA")
```

#### Filter SHS Telemetry to Desktop/Workstation Only
We use the final hardware dictionary as a strict whitelist to instantly purge all laptops, mobile chips, and integrated graphics from the market share data.

```{r}
shs_clean <- shs_clean %>%
  semi_join(hardware_dictionary, by = "chipset")

write_csv(shs_clean, "data/shs_clean.csv")
```

#### Export the Temporal Data
Use this to track pure market share and inflation trends over time

```{r}
time_series_data <- shs_clean %>%
  left_join(ppi_clean, by = "month_year")

write_csv(time_series_data, "data/time_series_temporal.csv", na = "NA")
```

#### Dataset Preparation for Modeling

Final refinements in preparation for the ISLR modeling phase: we calculate a dynamic inflation adjustment for GPU prices based on the PPI, which will allow us to analyze price trends in real terms over time.

```{r}
# Extract the absolute baseline PPI for our target month (December 2025)
baseline_ppi <- time_series_data %>%
  filter(month_year == ymd("2025-12-01")) %>%
  pull(semi_ppi) %>%
  unique()

final_modeling_dataset <- time_series_data %>%
  # Inner join inherently filters out laptop telemetry and merges specs
  inner_join(hardware_dictionary, by = "chipset") %>%
  mutate(
    # Dynamic Inflation Adjustment:
    # Calculates the real value of the GPU for that specific month in 2025 dollars
    inflation_adjusted_price = round(unified_price * (baseline_ppi / semi_ppi), 2)
  ) %>%
  select(
    month_year, chipset, percentage, semi_ppi,
    category, g3dmark, g2dmark, tensor_cores,
    unified_tdp, unified_memory,
    unified_price, inflation_adjusted_price
  ) %>%
  arrange(month_year, desc(percentage))

write_csv(hardware_dictionary, "data/hardware_dictionary_static.csv", na = "NA")
write_csv(hardware_dictionary_shs_only, "data/hardware_dictionary_shs_only.csv", na = "NA")
write_csv(final_modeling_dataset, "data/final_modeling_dataset.csv", na = "NA")
```

# Modeling Phase

To prevent severe overfitting, cross-validation is strictly required for model selection. It must be systematically deployed to:
1. Identify the precise tuning parameter $\lambda$ that minimizes test error in the Lasso models.
2. Select the optimal effective degrees of freedom (number of knots) for the Natural Cubic Splines within the GAM framework.

## Logistic Regression: Classification of Market Stratification

Model the conditional probability that a GPU is classified as an enterprise-grade unit based on its physical and architectural predictors.

$$\log\left(\frac{Pr(Y=\text{Workstation}|X)}{1-Pr(Y=\text{Workstation}|X)}\right)=\beta_0+\beta_1(\text{unified_memory})+\beta_2(\text{unified_tdp})+\beta_3(\text{tensor_cores})$$

Yielding large, positive, and statistically significant coefficients for specific AI predictors (e.g., tensor_cores) will empirically validate that high-end AI capabilities are systematically segmented and reserved for the workstation market.

```{r}
gpu_data <- read_csv("data/hardware_dictionary_static.csv")
# Data Preparation: Create the binary target variable
# We map 'Workstation' to 1 (Enterprise/Professional) and 'Desktop' to 0 (Consumer)
gpu_data$is_workstation <- as.factor(ifelse(gpu_data$category == "Workstation", 1, 0))

# Ensure predictors are numeric
gpu_data$unified_memory <- as.numeric(gpu_data$unified_memory)
gpu_data$unified_tdp <- as.numeric(gpu_data$unified_tdp)
gpu_data$tensor_cores <- as.numeric(gpu_data$tensor_cores)
gpu_data$g3dmark <- as.numeric(gpu_data$g3dmark)
gpu_data$g2dmark <- as.numeric(gpu_data$g2dmark)

# Subset and remove missing values so all models are trained on the exact same rows
model_vars <- c("is_workstation", "unified_memory", "unified_tdp", "tensor_cores", "g3dmark", "g2dmark")
model_data <- na.omit(gpu_data[, model_vars])

# Fit the Logistic Regression Model
# Model A: AI & Hardware Focused
model_ai <- glm(is_workstation ~ unified_memory + unified_tdp + tensor_cores,
                   data = model_data, family = binomial(link = "logit"))

# Output the Model Summary
# This will provide the coefficients (beta), standard errors, z-values, and p-values
summary(model_ai)
```
```{r}
# Calculate Odds Ratios for Interpretability
# Exponentiating the coefficients to understand the multiplicative effect on the odds
odds_ratios <- exp(coef(model_ai))
conf_intervals <- exp(confint(model_ai))

# Display Odds Ratios alongside 95% Confidence Intervals
cbind(Odds_Ratio = odds_ratios, conf_intervals)

# Optional: Predict Probabilities on the dataset to map the boundary
gpu_data$prob_workstation <- predict(model_ai, type = "response")
```

We look specifically at the Estimate and Pr(>|z|) (p-value) for tensor_cores and unified_memory.

* *Validating the AI Barrier*: If tensor_cores yields a large, positive, and statistically significant coefficient ($p < 0.05$), it empirically proves that high-end AI capabilities are systematically reserved for the workstation market.
* *Odds Ratios*: The exponentiated coefficient (Odds_Ratio) for unified_memory will tell you exactly how much the odds of a GPU being classed as a "Workstation" increase for every additional gigabyte of VRAM.

#### Comparing with other models

```{r}
# Model B: Traditional Rasterization Focused
model_raster <- glm(is_workstation ~ g3dmark + g2dmark,
                    data = model_data, family = binomial(link = "logit"))
# Model C: The Full Model (includes everything)
model_full <- glm(is_workstation ~ unified_memory + unified_tdp + tensor_cores + g3dmark + g2dmark,
                  data = model_data, family = binomial(link = "logit"))

# Automated Model Selection (Stepwise Algorithm)
# This systematically adds/removes variables to find the mathematically optimal model
# 'trace = 0' suppresses the wall of text it generates; set to 1 if you want to see the steps
model_auto <- step(model_full, direction = "both", trace = 0)
# View the final predictors chosen by the algorithm
summary(model_auto)
```

```{r}
# AIC (Akaike Information Criterion): Lower is better. It penalizes complex models.
aic_comparison <- AIC(model_ai, model_raster, model_auto)
# Calculate Delta AIC to see how far off the lesser models are from the best one
aic_comparison$Delta_AIC <- aic_comparison$AIC - min(aic_comparison$AIC)

cat("\n--- Model Comparison (Lowest AIC is Best) ---\n")
print(aic_comparison[order(aic_comparison$AIC), ])
# Likelihood Ratio Test
# Tests if the automated model provides a statistically significant improvement over your original AI model
cat("\n--- Likelihood Ratio Test: Original AI vs. Auto Selected ---\n")
anova(model_ai, model_auto, test = "Chisq")
```


## Lasso Regression: Regularization & Feature Selection

Resolve the severe multicollinearity existing between traditional gaming benchmarks and AI computational metrics within the high-dimensional feature space ($p$).

$$\min_{\beta}\left(\text{RSS}+\lambda\sum_{j=1}^p|\beta_j|\right)$$

The geometric properties of the $L_1$ norm penalty allow the Lasso to shrink the coefficient estimates of irrelevant or highly correlated predictors to exactly zero. If traditional gaming APIs (g3dmark) are penalized to zero while AI-specific metrics retain large positive weights, this provides rigorous econometric proof that the "AI Tax" is driving retail prices.

## Generalized Additive Models (GAMs): Non-Linear Market Economics

Model exponential scaling rules and physical boundaries by replacing rigid linear components with highly flexible, non-linear smoothing functions, $f_j(X_j)$.

$$\text{Price}=\beta_0+f_1(\text{g3dmark})+f_2(\text{tensor\_cores})+f_3(\text{FRED\_PPI})+\epsilon$$

#### Step Functions

Applied to VRAM capacity (unified_memory) to group continuous memory values into distinct bins, capturing the extreme price jumps that occur at specific architectural thresholds required to run specific AI models locally.

#### Natural Cubic Slines

Applied to continuous computational metrics (tensor_cores, FLOPs) to automatically map exponential price-to-performance curvature, avoiding erratic oscillation at the extremes of the data distribution.

