---
title: "AI Hardware Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Install required packages if missing:
# install.packages(c("tidyverse", "lubridate", "janitor"))

library(tidyverse)
library(lubridate)
library(stringr)
library(janitor)
```

# Data Processing

To accurately synthesize this data, a many-to-one relational join is required. This function standardizes names across all datasets to ensure clean joins.

```{r}
clean_gpu_string <- function(name_col) {
  name_col %>%
    str_to_lower() %>%
    # Remove brand fluff
    str_remove_all("msi|asus|gigabyte|zotac|evga|sapphire|powercolor|pny") %>%
    str_remove_all("nvidia|amd|ati|founders edition") %>%
    # Destroy hyphens followed by numbers (fixes the " - 512" issue)
    str_remove_all("-\\s*[0-9]+") %>%
    # Target VRAM (3gb), Wattage (75w), minor variants (g5, ddr4), and specific editions (se, oem, lhr, x2)
    str_remove_all("\\b([0-9]+gb|[0-9]+w|g[56]|ddr[0-9]|se|oem|lhr|x2|sdi)\\b") %>%
    # Remove the word "series" from SHS
    str_remove_all("\\bseries\\b") %>%
    # Squish multiple spaces down to a single space
    str_squish() %>%
    # Handle AMD Radeon HD groupings (e.g., changes "radeon hd 7750" to "radeon hd 7700")
    str_replace_all("radeon hd ([0-9]{2})[0-9]{2}[a-z]*", "radeon hd \\100") %>%
    # Handle AMD R7/R9 Series groupings (e.g., changes "r9 290x" to "r9 200")
    str_replace_all("r9 2[0-9]{2}[a-z]*", "r9 200") %>%
    str_replace_all("r9 3[0-9]{2}[a-z]*", "r9 300") %>%
    str_replace_all("r7 2[0-9]{2}[a-z]*", "r7 200") %>%
    str_replace_all("r7 3[0-9]{2}[a-z]*", "r7 300")
}
```

We load the five disparate datasets required for the high-dimensional dataset.
```{r}
# Assuming files are in the working directory
video_cards <- read_csv("data/video-card.csv") %>% clean_names()
gpu_benchmarks <- read_csv("data/GPU_benchmarks_v7.csv") %>% clean_names()
ml_hardware <- read_csv("data/ml_hardware.csv") %>% clean_names()
shs <- read_csv("data/shs.csv") %>% clean_names()
ppi_data <- read_csv("data/PCU334413334413.csv") %>% clean_names()
```

Process Retail Video Cards (video-card.csv)
Keeping NA/0 prices, dropping irrelevant string columns

```{r}
video_cards_clean <- video_cards %>%
  select(-name, -color, -length) %>%
  mutate(chipset = clean_gpu_string(chipset)) %>%
  group_by(chipset) %>%
  summarise(
    avg_retail_price = mean(price, na.rm = TRUE),
    vram_capacity = max(memory, na.rm = TRUE),
    core_clock = max(core_clock, na.rm = TRUE),
    boost_clock = max(boost_clock, na.rm = TRUE)
  ) %>%
  # Catch -Inf (from max) and NaN (from mean) and safely convert back to NA
  mutate(
    across(c(vram_capacity, core_clock, boost_clock), ~na_if(., -Inf)),
    avg_retail_price = na_if(avg_retail_price, NaN)
  )
```

Process Traditional Benchmarks (GPU_benchmarks_v7.csv)
Keeping price, dropping mobile segments

```{r}
benchmarks_clean <- gpu_benchmarks %>%
  mutate(
    # Create a temporary lowercase name for easy regex searching
    lower_name = str_to_lower(gpu_name),
    # Intelligently reclassify 'Unknown' (and correct mislabeled) GPUs
    category = case_when(
      # If it has laptop terminology, force it to Mobile
      str_detect(lower_name, "mobile|max-q|laptop|notebook") ~ "Mobile",
      # If it has enterprise/AI terminology, force it to Workstation
      str_detect(lower_name, "quadro|tesla|firepro|radeon pro|rtx a[0-9]") ~ "Workstation",
      # If it's still 'Unknown' after the above checks, safely assume it's a Desktop consumer card
      category == "Unknown" ~ "Desktop",
      # Otherwise, keep the original valid classification
      TRUE ~ category
    )
  ) %>%
  # Now apply the strict filter with confidence that the data is accurately tagged
  filter(category %in% c("Desktop", "Workstation")) %>%
  mutate(chipset = clean_gpu_string(gpu_name)) %>%
  select(chipset, category, g3dmark, g2dmark, tdp, benchmark_price = price) %>%
  distinct(chipset, .keep_all = TRUE)
```

Process Steam Hardware Survey (shs.csv)
Keeping 0 percentages, filtering out CPUs/OS metrics

```{r}
shs_clean <- shs %>%
  filter(category == "Video Card Description") %>%
  # Drop laptop chips (mobility/mobile) and integrated CPU graphics (intel)
  filter(!str_detect(str_to_lower(name), "mobility|mobile|intel")) %>%
  mutate(
    date = ymd(date),
    month_year = floor_date(date, "month"),
    chipset = clean_gpu_string(name)
  ) %>%
  # Apply strict timeline filter
  filter(month_year >= ymd("2009-01-01") & month_year <= ymd("2025-12-31")) %>%
  select(month_year, chipset, percentage)
```

Process AI Architectural Specs (ml_hardware.csv)
Filtering for GPUs only. Keeping all numeric specs, dropping irrelevant string columns

```{r}
ml_hardware_clean <- ml_hardware %>%
  filter(type == "GPU") %>%
  mutate(chipset = clean_gpu_string(hardware_name)) %>%
  select(
    chipset,
    ml_release_price = matches("release_price"),
    ml_memory = matches("memory_bytes"),
    memory_bandwidth = matches("memory_bandwidth"),
    ml_tdp = matches("tdp"),
    fp64_flops = matches("fp64_double"),
    fp32_flops = matches("fp32_single"),
    ml_base_clock = matches("base_clock"),
    ml_boost_clock = matches("boost_clock"),
    tensor_cores,
    transistors_millions,
    energy_efficiency
  ) %>%
  mutate(
    across(
      c(ml_memory, memory_bandwidth, fp64_flops, fp32_flops, energy_efficiency),
      ~ .x / 1e9
    )
  ) %>%
  distinct(chipset, .keep_all = TRUE)
```

Process Baseline Inflation Control (FRED's PCU334413334413.csv)
Filtering out historical data that precedes the SHS timeline

```{r}
min_shs_date <- min(shs_clean$month_year, na.rm = TRUE)
ppi_clean <- ppi_data %>%
  rename(date = observation_date, semi_ppi = pcu334413334413) %>%
  mutate(
    date = ymd(date),
    month_year = floor_date(date, "month")
  ) %>%
  # Apply strict timeline filter
  filter(month_year >= ymd("2009-01-01") & month_year <= ymd("2025-12-31")) %>%
  select(month_year, semi_ppi)
```

Write the Clean Datasets to CSV

```{r}
write_csv(video_cards_clean, "data/video_cards_clean.csv")
write_csv(benchmarks_clean, "data/benchmarks_clean.csv")
write_csv(shs_clean, "data/shs_clean.csv")
write_csv(ml_hardware_clean, "data/ml_hardware_clean.csv")
write_csv(ppi_clean, "data/ppi_clean.csv")
```

Export the Static Dictionary (No Repeated Rows)
Use this to easily see which exact GPUs are missing VRAM, FLOPs, or Prices

```{r}
hardware_dictionary <- video_cards_clean %>%
  full_join(benchmarks_clean, by = "chipset") %>%
  full_join(ml_hardware_clean, by = "chipset") %>%
  mutate(
    # Fixes NA categories for newer cards not in the benchmark dataset
    category = case_when(
      !is.na(category) ~ category, # Keep existing valid categories
      str_detect(chipset, "rtx|gtx|rx|radeon hd|arc a|arc b|geforce") ~ "Desktop", # Consumer gaming tags
      str_detect(chipset, "pro|quadro|tesla|instinct") ~ "Workstation", # Enterprise tags
      TRUE ~ "Workstation" # Default remaining unknowns to Workstation as requested
    ),
    # Convert all 0 values to NA across overlapping columns
    across(
      c(avg_retail_price, benchmark_price, ml_release_price,
        core_clock, ml_base_clock,
        boost_clock, ml_boost_clock,
        tdp, ml_tdp,
        vram_capacity, ml_memory),
      ~na_if(., 0)
    ),
    # Calculate Unified Means (suppressWarnings hides NaN warnings if all columns are NA)
    unified_price = suppressWarnings(rowMeans(cbind(avg_retail_price, benchmark_price, ml_release_price), na.rm = TRUE)),
    unified_base_clock = suppressWarnings(rowMeans(cbind(core_clock, ml_base_clock), na.rm = TRUE)),
    unified_boost_clock = suppressWarnings(rowMeans(cbind(boost_clock, ml_boost_clock), na.rm = TRUE)),
    unified_tdp = suppressWarnings(rowMeans(cbind(tdp, ml_tdp), na.rm = TRUE)),
    unified_memory = suppressWarnings(rowMeans(cbind(vram_capacity, ml_memory), na.rm = TRUE)),
    # Target Casting for Logistic Regression
    is_workstation = as.factor(ifelse(category == "Workstation", 1, 0)),
    # Ensure categorical features are appropriately typed
    category = as.factor(category)
  ) %>%
  # Convert any NaN (caused by averaging all-NA columns) back to proper NA for clean CSV export
  mutate(across(starts_with("unified_"), ~na_if(., NaN))) %>%
  # Drop the original messy columns
  select(
    -avg_retail_price, -benchmark_price, -ml_release_price,
    -core_clock, -ml_base_clock,
    -boost_clock, -ml_boost_clock,
    -tdp, -ml_tdp,
    -vram_capacity, -ml_memory
  ) %>%
  drop_na(unified_price)

write_csv(hardware_dictionary, "data/hardware_dictionary_static.csv", na = "NA")

hardware_dictionary_shs_only <- hardware_dictionary %>%
  semi_join(shs_clean, by = "chipset")

write_csv(hardware_dictionary_shs_only, "data/hardware_dictionary_shs_only.csv", na = "NA")
```

Export the Temporal Data
Use this to track pure market share and inflation trends over time

```{r}
time_series_data <- shs_clean %>%
  left_join(ppi_clean, by = "month_year")

write_csv(time_series_data, "data/time_series_temporal.csv", na = "NA")
```

Export the Unified Dataset (STRICTLY FOR MODELING)

```{r}
final_analytical_dataset <- time_series_data %>%
  inner_join(hardware_dictionary, by = "chipset") %>%
  mutate(
    # Create the baseline inflation control target
    inflation_adjusted_price = unified_price / (semi_ppi / 100)
  )

write_csv(final_analytical_dataset, "data/final_analytical_dataset.csv", na = "NA")
```

